\chapter*{Acknowledgements}
I would like to thank my PhD supervisor Prof. Augusto Sarti for his suggestions and guidance throughout my PhD. I am also grateful to Prof. Stefano Tubaro for his advice. I would like to thank Massimiliano Zanoni, who has co-supervised my PhD for supporting me during the tough moments and some climbing occasions.

I would also like to thank Prof. Mark Sandler and Prof. George Fazekas of the Queen Mary University of London, for their supervision during my period at the Centre for Digital Music laboratory (C4DM). The opportunity of experimenting a different research environment has greatly enriched my experience. I am also grateful to all the C4DM people who host me, with a special mention for Mi, Keanwoi, Florian, Delia, Alo, Sebastian, Adip and Thomas. 

I am grateful to the thesis reviewers, Prof. Slim Essid and Prof. Mathieu Barthet for the effort they dedicated in the review of my thesis work, which ultimately helped me improve it.

I must thank all the ISPG people for the great work environment they have created: Fabio Antonacci, Alberto Bernardini, Paolo Bestagini, Lucio Bianchi, Luca Bondi, Federico Borra, Antonio Canclini, Bruno Di Giorgi, Silvia Lameri,  Sara Mandelli, Marco Marcon, Dejan Markovic,  Simone Milani, Ambra Melloni,  Davide Sanvito,  Francesco Setragno, Muhammad Shahnawaz, Marco Tagliasacchi, the students I had the honor to co-advise, Davide Andreoletti and Alessandro Gallo, and the guests Filipe Costa and Siva Prasad Nandyala.
I must also thank Francesca Clemenza, who helped me greatly during for the organization of the burocracy.  

A special thank to my family: my parents Cesarina and Gigi, and my siblings Carlo, Laura and Beatrice, for their precious ability to tease me.

I would like to mention my family in Milan: my friends Angela, Rocco, Amanda, Pietro, Ilaria, Antonio and Marianna, for their lovely support and fun. 

Finally, I must thank Francesca, because, as Han Solo once said, \textit{I know}.

%Last but not least, I have to thank Francesca, for her long-time \textit{wonderfulness}.%, for having been there, for being there, for being her.  

